---
title: "A quick <br/ >(*equation free*)<br /> intro to Hidden Markov Models applied to Stock Volatility"
author: "Luis Damiano"
date: "R/Finance 2017 | May 19"
output: 
  revealjs::revealjs_presentation:
    bibliography: ../references.bib
    incremental: false
    css: custom.css
    theme: sky
    highlight: tango
    center: false
    smart: true
    transition: zoom
    self_contained: false
    lib_dir: lib
    reveal_plugins: ["notes"]
    reveal_options:
      controls: true
      progress: true
      keyboard: true
      center: false
      hideAddressBar: true
      slideNumber: true
      previewLinks: true
      fig_width: 7
      fig_height: 6
      fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

<!--
A few basic questions that should serve as a guide while making the slides:

* What does "hidden" mean
* What does markov mean in the context of an hmm
* Where are some obvious fits for this in trading and finance
* A financial time series example

# Agenda

## Agenda

- Motivation (30')
- Hidden Markov Models (60')
- Application to stock vol (180')
- Takeaway (30')
-->

# Motivation

## Ever heard this question?
<span class="bigtext">How's the <span class="largetext mygreen">market</span> today?</span>

## Possible answers
<aside class="notes">
- Don't rush thru this slide - try to catch their attention
- The answer depends on what you're modelling, but they all have something in common...
</aside>

- **Trader**: bullish/bearish
- **Risk analyst**: low/high volatility
- **Behavioural finance**: risk on/off
- **Macro economist**: expansion/recession

## How's the market today?
<aside class="notes">
- An assumption everyone in this room will easily buy
- Market is too complex to be handled by a simple model that holds in every single context
- Otherwise, we'd be dead bored at work
- ex. Different models for low/high vol or bull/bear markets
</aside>

This *fair* question has an *implicit* assumption:

<br />

Markets <br />
<span class="bigtext myred">DON'T</span><br />
behave<br />
<span class="largetext">*exactly*</span><br />
the same way<br />
every single day.

## Markets have states instead
But they're<br />
<span class="bigtext">**HIDDEN**</span><br />

>- can't be observed directly
>- need inference based on observed data (prices, volume, order size)

# Hidden Markov Models {data-background="./img/hmm.svg" data-background-size="600px" data-background-repeat="no-repeat"}

## Specification

<aside class="notes">
- don't explain the keywords!
- re Box: remember models are simplifications of reality.
- btw, Box was an awesome statistician
</aside>

HMM involves two interconnected models

>- **State model**: discrete-time, discrete-state Markov chain with hidden states.
>- **Observation model**: a model for the observations, emissions or output.
>- **Keywords**: bayesian inference, probabilistic machine learning, state space models.

> All models are wrong, but some are useful. (George E. P. Box)

## HMM and Mixtures

<aside class="notes">
- Components have memory
</aside>

> HMM may be interpreted as an extension of a mixture model

>- Each single time slice corresponds to a mixture, but...
>- The component for each observation has memory.

# HMM applied to stock vol!

<!-- Pic here -->

```{r message = FALSE, dpi=300, warning = FALSE}
library(quantmod)
syms <- c('^GSPC', 'F', 'GM', 'THO', 'AIR')

usecache = TRUE
cache.filename <- 'data/symsret.RDS'
if (usecache && file.exists(cache.filename)) {
  y_t <- readRDS(cache.filename)
} else {
  y_t = as.matrix(
    do.call(cbind, lapply(syms, function(s) {
      p <- getSymbols(s, src = "yahoo", from = "2011-01-01", to = "2016-12-31", auto.assign = FALSE)
      r <- periodReturn(p, period = 'daily', type = 'log')
      colnames(r) <- s
      r * 100
    })))
}

library(rstan)
rstan_options(auto_write = TRUE)  # Writes down the compiled sampler
options(mc.cores = parallel::detectCores()) # Use all available cores

K = 2 # Number of states
bmodel = 'stan/hmmgarch.stan' # Name of the model file
standata = list(
  N = ncol(y_t),
  T = nrow(y_t),
  K = K,
  y = t(y_t)
)

usecache = TRUE
cache.filename <- paste0('stan/hmmgarch-cache.rds')
if (usecache && file.exists(cache.filename)) {
  stan.fit <- readRDS(cache.filename)
} else {
  stan.fit <- stan(file = bmodel,
                   model_name = "Bayesian Hierarchical Mixture GARCH",
                   data = standata, verbose = T,
                   iter = 200, warmup = 100, thin = 1, chains = 4, cores = 4,
                   control = list(adapt_delta = 0.80))
  
  saveRDS(stan.fit, file = cache.filename)
}

library(gridBase)
library(gridExtra)
library(lattice)
library(HH)

cols = c("#E41F26", "#2EA147", "#1D79B4", "#fdb863", "#E6E6E6", "#cfcfcf", "#63656A")

colMedians <- function(x) {
  f <- median # You can switch to 'mean' if you dare...
  
  if (length(dim(x)) == 4) {
    apply(x, c(1, 4), function(y) { f(y) })
  } else if (length(dim(x)) == 3) {
    apply(x, c(2, 3), function(y) { f(y) })
  } else if (length(dim(x)) == 2) {
    apply(x, 2, function(y) { f(y) } )
  } else {
    f(x)
  }
}
```

## Setup

>- **Goal**: modeling deterministic volatility for stocks
>- **Data**: Daily close prices 2011-2016 from ^GSPC, F, GM, THO, AIR.
>- **State model**: binary latent state with first-order markovian memory
>- **Observation model**: $GARCH(1, 1)$

## One stock (F) with two latent states

<aside class="notes">
- Explain the vertical axis in plot 2.
</aside>

```{r, fig.width = 18, fig.height = 6, fig.keep = 'last'}
  n <- 2 # Ford
  library(grid)
  # Time series
  tsmat <- cbind(
    colMedians(extract(stan.fit, pars = 'sigma_t')[[1]][, n, 1, ]),
    colMedians(extract(stan.fit, pars = 'sigma_t')[[1]][, n, 2, ]))
  colnames(tsmat) <- c('Low state', 'High state')
  
  tsplot <- xyplot.ts(tsmat,
    # main = 'Conditional volatility - GARCH(1, 1)',
    ylab = bquote(sigma[t]), xlab = bquote(t),
    superpose = TRUE, outside = TRUE, ylim = c(0, max(tsmat)),
    scales = list(cex = 2, tck = 0.5),
    col = cols[c(2, 1)],
    key = list(
      text = list(colnames(tsmat)), space = 'top',
      columns = ncol(tsmat), corner = c(0.5, 0.95),
      cex = 2, lines = list(lwd = 2, size = 2, col = cols[c(2, 1)]),
      between = 0.2
    ))

  # Cross sectional
  csmat <- as.data.frame(
    cbind(
      x = t(colMedians(extract(stan.fit, pars = 'F')[[1]][, , ]))[, n], 
      y = as.vector(y_t[, n])))
  
  cpplot <- xyplot(x ~ y, data = csmat,
         ylim = c(0, 1),
         scales = list(cex = 2, tck = 0.5),
         ylab = bquote(pi[1 * t]),
         xlab = bquote(r[t]),
         # main = 'Logreturns vs. Low volatility state',
         panel = function(x, y, ...) {
           panel.xyplot(x, y, pch = 21, col = cols[6], bg = cols[5], cex = 0.75, ...)
           panel.loess(x, y, span = 2/3, degree = 1, family = 'gaussian', lty = 1, lwd = 2, col = cols[2])
         },
        key = list(
          text = list(c('Loess')), space = 'top',
          columns = 1, corner = c(0.5, 0.95),
          cex = 2, lines = list(lwd = 2, size = 2, col = cols[c(2)]),
          between = 0.2
    ))
  
  # Plot
  grid.arrange(tsplot, cpplot,
               layout_matrix = matrix(
                 1:2, byrow = TRUE, ncol = 2, nrow = 1)
  )
```

<span class = "smalltext">**Common sense check**: Observing small returns makes us more confindent about using the low volatility model.</span>

<!-- ## Other characteristics (see the appendix) -->
<!-- - State probabilities are autocorrelated, hinting that some kind of memory structure may help in prediction (not shown). -->
<!-- - The variance of the returns in each state are linearly related (not shown). -->
<!-- - You should test if these features are reasonable for your own use case! -->

## A stock portfolio with two latent states
```{r, fig.width = 18, fig.height = 6, fig.keep = 'last'}
  mat <- t(colMedians(extract(stan.fit, pars = 'F')[[1]][, , ]))
  colnames(mat) <- syms
  n <- c(2, 3, 5)
  
  csplot <- function(x, y, ...) {
    xyplot(x ~ y, data = data.frame(x = x, y = y),
          panel = function(x, y, ...) {
            panel.xyplot(x, y, pch = 21, col = cols[6], bg = cols[5],
                         cex = 0.5, xlim = c(0, 1), ylim = c(0, 1), ...)
            panel.abline(lm(y ~ x), lty = 1, lwd = 2, col = cols[4], ...)
            panel.loess(x, y, span = 1/3, degree = 1, family = 'gaussian',
                        lty = 1, lwd = 2, col = cols[2])
            panel.text(0.5, 0.5,  cex = 2,
                       eval(parse(text = paste('bquote(hat(rho) == .(sprintf(\'%0.2f\', ', cor(x, y), ')))')))
            )
          },
          scales = list(cex = 2, tck = 0.5),
          key = list(
            text = list(c('Linear', 'Loess')), 
            columns = 2, lines = list(col = cols[c(4, 2)]), 
            space = 'top', cex = 2, lwd = 2), ...)}

  csplot1 <- csplot(mat[, 3], mat[, 2], xlab = 'GM', ylab = 'F')
  csplot2 <- csplot(mat[, 5], mat[, 2], xlab = 'AIR', ylab = 'F')

  # Ready, steady, plot!
  grid.arrange(csplot1, csplot2,
               layout_matrix = matrix(
                 1:2, byrow = TRUE, ncol = 2, nrow = 1)
  )
```

<span class = "smalltext">Correlation in belief states is higher for stocks in the same industry (compare F/GM vs F/AIR).</span>

## Takeaway
<aside class="notes">
- Explain the vertical axis in plot 2.
- Say, the prob that similar stocks are in the high volatility state are correlated.
</aside>

>- States are hidden, yet meaningful. They are subject to useful domain knowledge interpretation.
>- Belief states across assets are cross-correlated.
>- This relationship is stronger for stocks in the same industry, with similar business models or similar exposure to macroeconomic factors.

## Further research

> What if hidden states were **hierarchical**?

<span class="smalltext">Risk state of a global portfolio may be mapped into components</span>
<br />$\text{Country} + \text{Industry} + \text{Individual}$

I'm currently exploring <br />**Hierarchical Hidden Markov Models applied to finance**<br /> for my Google Summer of Code 2017 project.

## Thanks! {.gitlogo}

<h2>May any of these ideas be of value for your trading strategy or risk model?</h2>

<span class="smalltext">Check the extended material for this talk
<br />(R Notebook, fully working code, references and neat plots)</span>

[https://github.com/luisdamiano/rfinance17](https://github.com/luisdamiano/rfinance17)

![2](img/1493974219_github.svg)
